---
title: "Winter 2020 BSA Article"
author: 'Ethan Allavarpu and Kyle Boal'
date: "1/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = NA)
```

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

<!-- # Gather and Clean Data -->
<!-- ## Function for Sorting Games by Date -->
```{r}
ordered_dates <- function(date_list) {
  for (i in 1:2) {
    if (nchar(date_list[i]) < 2) {
      date_list[i] <- paste(0, date_list[i],
                            sep = "",
                            collapse = "")
    }
  }
  order_date <- paste(date_list[3], date_list[1], date_list[2],
                      sep = "",
                      collapse = "")
  as.numeric(order_date)
}
```
<!-- ## Create .csv Files for Exportation -->
<!-- ### 2015 - 2019 -->
```{r}
cointoss20152019 <- read.csv("../data/cointoss2015_2019.csv", header = TRUE, stringsAsFactors = TRUE)
names(cointoss20152019)[2] <- "Team Won Toss"
cointoss20152019 <- cointoss20152019[-1][-2][-3][-3][-3][-4][-4][-4][-5][1:5]
cointoss20152019 <- cointoss20152019[1:54, ]
results <- as.character(cointoss20152019$Result)
results <- substring(results, 1, 1)
results <- factor(results)
cointoss20152019$Result <- results
dates <- as.character(cointoss20152019$Date)
dates_list <- strsplit(dates, "/")
dates_list <- lapply(dates_list, as.numeric)
date_order <- vapply(dates_list, ordered_dates, numeric(1))
names(date_order) <- 1:54
date_sorted <- sort(date_order)
cointoss20152019 <- cointoss20152019[names(date_sorted),]
# write.csv(cointoss20152019, "../data/20152019cointoss.csv")
```

<!-- ### 2002 - 2006 -->
```{r}
cointoss20022006 <- read.csv("../data/cointoss2002_2006.csv", header = TRUE, stringsAsFactors = TRUE)
names(cointoss20022006)[2] <- "Team Won Toss"
cointoss20022006 <- cointoss20022006[-1][-2][-3][-3][-3][-4][-4][-4][-5][1:5]
cointoss20022006 <- cointoss20022006[1:55, ]
results <- as.character(cointoss20022006$Result)
results <- substring(results, 1, 1)
results <- factor(results)
cointoss20022006$Result <- results
dates <- as.character(cointoss20022006$Date)
dates_list <- strsplit(dates, "/")
dates_list <- lapply(dates_list, as.numeric)
date_order <- vapply(dates_list, ordered_dates, numeric(1))
names(date_order) <- 1:55
date_sorted <- sort(date_order)
cointoss20022006 <- cointoss20022006[names(date_sorted),]
# write.csv(cointoss20022006, "../data/20022006cointoss.csv")
```

<!-- ### 2007 - 2014 -->
```{r}
cointoss20072014 <- read.csv("../data/cointoss2007_2014.csv", header = TRUE, stringsAsFactors = TRUE)
names(cointoss20072014)[2] <- "Team Won Toss"
cointoss20072014 <- cointoss20072014[1:88, ]
cointoss20072014 <- cointoss20072014[-1][-2][-3][-3][-3][-4][-4][-4][-5][1:5]
results <- as.character(cointoss20072014$Result)
results <- substring(results, 1, 1)
results <- factor(results)
cointoss20072014$Result <- results
# write.csv(cointoss20072014, "../data/20072014cointoss.csv")
```

# Redownloading the Data
*Notes:*  

  1. Spread is given for the team that won the toss. A positive spread indicates an underdog, a negative spread indicates a favorite.  
  2. For coin toss choice, "R" means that the team chose to receive, while "D" means that the team chose to defer.
  
## 2015 - 2019
```{r}
cointoss20152019 <- read.csv("../data/20152019cointoss.csv", header = TRUE, row.names = 1, stringsAsFactors = TRUE)
cointoss20152019 <- cbind(cointoss20152019, "Year" = rep(2015:2019, each = 11))
knitr::kable(cointoss20152019, align = "c", row.names = FALSE)
```

## 2002 - 2006
```{r}
cointoss20022006 <- read.csv("../data/20022006cointoss.csv", header = TRUE, row.names = 1, stringsAsFactors = TRUE)
cointoss20022006 <- cointoss20022006[, 1:6]
cointoss20022006 <- cbind(cointoss20022006, "Year" = rep(2002:2006, each = 11))
knitr::kable(cointoss20022006, align = "c", row.names = FALSE)
```

```{r}
cointoss20072014 <- read.csv("../data/20072014cointoss.csv", header = TRUE, stringsAsFactors = TRUE)
cointoss20072014 <- cointoss20072014[seq(from = 2,
                                         to = nrow(cointoss20072014),
                                         by = 2), -1]
cointoss20072014 <- cbind(cointoss20072014, "Year" = rep(2007:2014, each = 11))
names(cointoss20072014) <- names(cointoss20152019)
knitr::kable(cointoss20072014, align = "c")
```


# Data Analysis
## Is there a difference in deferral percentage across eras?  
$H_0$: $p_1 = p_2$  
$H_a$: $p_1 < p_2$  

  +  $p_1$: Deferral percentage for 2002 - 2006  
  +  $p_2$: Deferral percentage for 2015 - 2019  
  
```{r}
X1 <- sum(cointoss20022006$Coin.Toss.Choice == "D")
X2 <- sum(cointoss20152019$Coin.Toss.Choice == "D")
p1 <- mean(cointoss20022006$Coin.Toss.Choice == "D")
p2 <- mean(cointoss20152019$Coin.Toss.Choice == "D")
barplot(c("2002 - 2006" = p1, "2015 - 2019" = p2),
        main = "Percent of Coin Tosses with Deferral Result",
        xlab = "Year Span",
        ylab = "Percent Deferrals",
        col = c("red", "blue"),
        density = 100 * c(p1, p2),
        las = 1)
text(0.7, p1 + 0.05, labels = round(p1, 3))
text(1.9, p2 - 0.05, labels = round(p2, 3), col = "white")
def_prop <- prop.test(c(X1, X2),
                      c(nrow(cointoss20022006), nrow(cointoss20152019)),
                      alternative = "less",
                      correct = FALSE)
def_prop
p_val <- def_prop$p.value
p_val
```
Since $p < \alpha$, we reject $H_0$. We do have evidence which supports the alternative hypothesis $H_a$ that the deferral percentage for winners of the coin toss is less for the years 2002 - 2006 than for the years 2015 - 2019.  

## Does deferring provide a statistically significant advantage in win percentage?
$H_0$: $p_1 = p_2$  
$H_a$: $p_1 \neq p_2$  

  +  $p_1$: Win percentage for deferring
  +  $p_2$: Win percentage for receiving

```{r}
cointoss <- rbind(cointoss20022006, cointoss20072014, cointoss20152019)
defer <- cointoss[cointoss$Coin.Toss.Choice == "D", ]
receive <- cointoss[cointoss$Coin.Toss.Choice == "R", ]
defer_wins <- sum(defer$Result == "W")
receive_wins <- sum(receive$Result == "W")
defer_win_pct <- mean(defer$Result == "W")
receive_win_pct <- mean(receive$Result == "W")
barplot(c("Defer" = defer_win_pct, "Receive" = receive_win_pct),
        main = "Win Percentage by Coin Toss Decision",
        xlab = "Coin Toss Decision",
        ylim = c(0, 0.6),
        ylab = "Win Percentage",
        col = c("red", "blue"),
        density = 100 * c(defer_win_pct, receive_win_pct),
        las = 1)
text(0.7, defer_win_pct + 0.05,
     labels = paste(round(defer_win_pct, 3), " (",
                    defer_wins, " instances)", sep = ""))
text(1.9, receive_win_pct + 0.05,
     labels = paste(round(receive_win_pct, 3), " (",
                    receive_wins, " instances)", sep = ""))
win_prop <- prop.test(c(defer_wins, receive_wins),
                      c(nrow(defer), nrow(receive)),
                      correct = FALSE)
win_prop
p_val <- win_prop$p.value
p_val
```


```{r}
ctwin <- mean(cointoss$Result == "W")
ctlose <- 1 - ctwin
barplot(c("Won Coin Toss" = ctwin, "Lost Coin Toss" = ctlose),
        main = "Win Percentage for Teams Winning or Losing the Coin Toss",
        xlab = "Won Coin Toss?",
        ylim = c(0, 0.7),
        ylab = "Win Percentage",
        col = c("red", "blue"),
        density = 100 * c(ctwin, ctlose),
        las = 1)
text(0.7, ctwin + 0.05,
     labels = round(ctwin, 3))
text(1.9, ctlose + 0.05,
     labels = round(ctlose, 3))
```


# Data Visualization
## Line Plots
```{r}
receive_pct <- tapply(as.numeric(cointoss$Coin.Toss.Choice),
                      cointoss$Year,
                      mean) - 1
defer_pct <- 1 - receive_pct
plot(unique(cointoss$Year), defer_pct, col = "blue", pch = 17,
     cex = 0.75, ylim = c(0, 1),
     type = "o",
     ylab = "Percent",
     xlab = "Year",
     main = "Deferral Percentage and\nWin Percentage for Teams that Deferred")
defers <- cointoss[cointoss$Coin.Toss.Choice == "D", ]
defer_win_pct_year <- tapply(as.numeric(defers$Result),
                             defers$Year,
                             mean) - 1
lines(unique(defers$Year), defer_win_pct_year, col = "red",
       lty = 2, type = "o", pch = 19,
       cex = 0.75)
year <- unique(cointoss$Year)
abline(h = 0.5, col = "black", lty = 2)
legend("left",
       legend = c("Defer Win Percentage",
                  "Defer Percentage"),
       pch = 19,
       cex = 0.75,
       col = c("red", "blue"),
       inset = 0.05)

plot(unique(cointoss$Year), receive_pct, col = "blue", pch = 17,
     cex = 0.75, ylim = c(0, 1),
     type = "o",
     ylab = "Percent",
     xlab = "Year",
     main = "Receive Percentage and\nWin Percentage for Teams that Received")
receives <- cointoss[cointoss$Coin.Toss.Choice == "R", ]
rec_win_pct_year <- tapply(as.numeric(receives$Result),
                             receives$Year,
                             mean) - 1
lines(unique(receives$Year), rec_win_pct_year, col = "red",
       lty = 2, type = "o", pch = 19,
       cex = 0.75)
abline(h = 0.5, col = "black", lty = 2)
legend("bottomleft",
       legend = c("Receive Win Percentage",
                  "Receive Percentage"),
       pch = 19,
       cex = 0.75,
       col = c("red", "blue"),
       inset = 0.05)
```

## Deferrals by Round
```{r, out.width="100%"}
colors <- c("coral1",
            "darkorchid3",
            "deepskyblue",
            "khaki")
cointoss$Round <- factor(c(rep(c(rep(c("1. W", "2. D"), each = 4),
                          "3. C", "3. C", "4. SB"), 18)),ordered = TRUE)
round_receive_pct <- tapply(as.numeric(cointoss$Coin.Toss.Choice),
                      list(cointoss$Year, cointoss$Round),
                      mean) - 1
round_defer_pct <- 1 - round_receive_pct
barplot(t(round_defer_pct), beside = TRUE, col = colors, las = 1,
        main = "Deferral Percentage by Year and Round",
        xlab = "Year",
        ylab = "Deferral Percentage")
legend("topleft",
       title = "Round",
       legend = c("Wild Card",
                  "Divisional ",
                  "Conference Championship",
                  "Super Bowl"),
       col = colors,
       lty = 1,
       lwd = 1,
       cex = 0.5,
       inset = 0.05)
```
```{r}
par(mfrow = c(2,2))
rounds <- c("Wild Card", "Divisional",
            "Conference Championship", "Super Bowl")
for (i in 1:4) {
  barplot(round_defer_pct[, i], col = colors[i], las = 1,
        main = rounds[i],
        xlab = "Year",
        ylab = "Deferral Percentage")
}
```

```{r}
rounds_receive_pct <- tapply(as.numeric(cointoss$Coin.Toss.Choice),
                             cointoss$Round,
                             mean) - 1
rounds_defer_pct <- 1 - rounds_receive_pct
names(rounds_defer_pct) <- c("Wild Card", "Divisional", "Championship",
                             "Super Bowl")
barplot(rounds_defer_pct, col = colors, ylab = "Deferral Percentage",
        las = 1, main = "Deferral Percentage by Round")
```

## Four Sample Test for Equality of Proportions (Chi-Square GoF)
```{r}
rounds_defer_number <- c(rounds_defer_pct[1:2] * 4,
                         rounds_defer_pct[3] * 2,
                         rounds_defer_pct[4])
rounds_defer_number <- 18 * rounds_defer_number
prop.test(rounds_defer_number, 18 * c(4, 4, 2, 1))
```

## Chi Square Test for Independence
*Test if the proportion of deferrals in each round remains the same over all years*
```{r}
defers <- defer_pct * 11
defers <- defers[defers > 0]
prop.test(defer_win_pct_year * defers, defers)
```

## Relationship between Defer Percentage and Win Percentage
```{r}
defer_pct
defer_win_pct_year
defer_win_pct_year <- c(defer_win_pct_year[1:3], 0, 0, defer_win_pct_year[4:length(defer_win_pct_year)])
defer_win_pct_year
plot(defer_win_pct_year ~ defer_pct,
     pch = 19,
     col = rgb(0, 0, 0, alpha = 0.25))
win_to_defer <- lm(defer_win_pct_year ~ defer_pct)
summary(win_to_defer)
```


